Applications of the cross-entropy method to importance sampling and optimal control of diffusions

### 2014-04-29 writedown/first.hitting/cross_entropy/cross_entropy.tex
update the data
backup

\newcommand{\rates}[2]{\displaystyle \mathrel{\longrightleftharpoons^{#1\mathstrut}_{#2}}}
\newcommand{\cL}{{\mathcal L}}
\newcommand{\bR}{{\mathbb R}}
\newcommand{\argmin}{\mathop{\rm argmin}}%

\newcommand{\bE}{{\mathbf E}}

@@

\begin{document}
\title{Applications of the cross-entropy method in diffusion processes}
%\author{Wei Zhang}
\author{Wei Zhang\textsuperscript{2}\and Han Wang\textsuperscript{1} \and
  Carsten Hartmann\textsuperscript{1,*}\and 
Marcus Weber\textsuperscript{2}\and Christof Sch\"{u}tte\textsuperscript{1,2}}
\footnotetext[1]{Institute of Mathematics, Freie Universit\"{a}t Berlin, Arnimallee 6, 14195 Berlin, Germany}
\footnotetext[2]{Zuse Institute Berlin, Takustrasse 7, 14195 Berlin, Germany}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[1]{Corresponding author. Email: chartman@mi.fu-berlin.de}
\date{}
\maketitle
\begin{abstract}
  The cross-entropy method in the diffusion process setting is studied. The
  general algorithm is proposed and is applied to design efficient importance sampling for rare events as well as to solve optimal
  control problem of diffusion process. Several examples are studied
  numerically to demonstrate this method.
\end{abstract}
\begin{keywords}
importance sampling, optimal control, cross-entropy method, Monte Carlo
method, rare event, change of measure. 
\end{keywords}
\begin{AMS}
\end{AMS}
\section{Introduction}
\label{intro}
Diffusion process has been extensively investigated because of its wide
applications in mathematics, physics, economics, and so on.
Here we are mainly interested in two topics in diffusion process, i.e. importance sampling and optimal control. 
For importance sampling, it is one kind of variance reduction technique 
in Monte Carlo simulations and becomes especially necessary in rare event simulation.
For optimal control, the purpose is to driven the dynamical system in certain
optimal way and has strong applicational backgrounds in finance, climate
science, mechanics \cite{fleming2006}. We emphasize that these two different areas are actually related and can
be unified as the same problem, i.e. how to change the measure in certain optimal way.

Here we take an applied point of view, i.e. we aim at designing efficient
algorithms to find the way to do the importance sampling, or to find the
optimal control strategy. One main reason that designing efficient algorithms
is urgent and challenging is due to the high dimensionality of the dynamical
systems, as in the case of climate system, biological system and so on. However, when the dynamical
systems have multiscale feature, efficient algorithms may be designed by
taking this feature into account. See \cite{ip-dupuis-multiscale, ip-eric,
zhws13} on importance sampling and \cite{control_schuette, zlph2013} on
optimal control for more details.

In this work, we consider these two topics together as how to design the change of
measure in an optimal way. If we further assume the existence of the optimal
change of measure $\mu^*$, the problem can be roughly described as how to find the approximation of
$\mu^*$ under some measure distance. For the choice of measure distances, we will use the Kullback-Leibler
divergence, which, although not a metric, is
widely used to quantified the similarity of two probability measures. 
What is more, given certain target probability measure $\mu^*$, the so-called cross-entropy method has been
developed recently and provides a general algorithm to find the minimizer of the
Kullback-Leibler divergence w.r.t. $\mu^*$, among a family of parameterized
probability measures. It has wide applications in different areas, such as
Monte Carlo simulation, optimization, and machine learning \cite{ce_tutorial, ce_book}. The
main purpose of this paper is to formulate the cross-entropy method in the
context of diffusion process, i.e. for measures defined in path space, and
discuss the applications of this method in importance sampling and optimal control
problems.

This paper is organized as follows. In Section~\ref{sec-om-functional},
using Girsanov theorem, we derive the Onsager-Machlup functional for a given trajectory.
In Section~\ref{sec-ce-is}, the cross-entropy method in path space is
formulated for the importance sampling problem.
In Section~\ref{sec-ocp}, by using the dual relationship between importance
sampling and optimal control problem, we discuss the cross-entropy method for
optimal control problem. The cross-entropy algorithm is summarized in
Section~\ref{sec-ce-algo}. Several numerical examples are studied in
Section~\ref{sec-examples}. Discussions are presented in
Section~\ref{sec-discuss}.

\section{Onsager-Machlup functional}
\label{sec-om-functional}
Let $O \subset \mathbb{R}^n$, $z_s \in \mathbb{R}^n$ satisfies the SDE 
\begin{align}
\begin{split}
  d z_s &= b(z_s) ds + \sqrt{2\beta^{-1}} dw_s, \quad 0 < s < \tau  \\
  z_0&=x
  \end{split}
\label{dynamics-1}
\end{align}
$\beta >0$ is constant, $\tau = \inf_{s>0} \{s;\, z_s \in O\}$ is some
stopping time, $b : \mathbb{R}^n \rightarrow \mathbb{R}^n$, $w$ is a
$n$-dimensional Brownian motion.

We derive the Onsager-Machlup functional for dynamics (\ref{dynamics-1}) in
this section.
Let $\mu$ and $\mu_0$ be the probability measures on path space generated by
(\ref{dynamics-1}) and the scaled Brownian motion
\begin{align}
  \begin{split}
  dz_s &= \sqrt{2\beta^{-1}}dw_s, \hspace{3mm} 0 < s < \tau \\
   z_0 &= x .
  \end{split}
From Girsanov theorem \cite{oksendalSDE}, we have 
\begin{align}
  \frac{d\mu}{d\mu_0} &= \exp\Big(\sqrt{\frac{\beta}{2}}\int_0^\tau
  \langle b(z_s), dw_s\rangle + \frac{\beta}{4}\int_0^\tau |b(z_s)|^2ds\Big) \notag\\
  &=\exp\Big(\frac{\beta}{2}\int_0^\tau \langle b(z_s),dz_s\rangle -
  \frac{\beta}{4}\int_0^\tau |b(z_s)|^2ds\Big) 
and
\begin{align}
  d\mu_0 = \exp\Big(-\frac{\beta}{4}\int_0^\tau
  \Big|\frac{dz}{ds}\Big|^2ds\Big) d\nu\,
where $\nu$ is the Wiener measure. Together we obtain
\begin{align}
  d\mu &=\exp\Big(-\frac{\beta}{2} I_{1}(z)\Big)d\nu
  \label{measure-1}
where
\begin{align}
  I_{1}(z) = \frac{1}{2}\int_0^\tau \Big|\frac{dz}{ds} - b(z_s)\Big|^2 ds
and the integral is interpreted as Ito integral.
We could also rewrite it as Stratonovich integral using the relationship
\begin{align}
  \int_0^\tau \langle b(z_s), dz_s\rangle = \int_0^\tau b(z_s) \circ dz_s - \beta^{-1} \int_0^\tau \operatorname{div}\,b(z_s) ds
and therefore obtain 
\begin{align}
  d\mu = \exp \Big(-\frac{\beta}{2} I_2(z) \Big) d\nu\,
where functional 
\begin{align}
  I_2(z) = \frac{1}{2}\int_0^\tau \Big|\frac{dz}{ds}- b(z_s)\Big|^2 ds +
  \beta^{-1} \int_0^\tau \operatorname{div}\,b(z_s) ds
  \label{om-fun}
is often called Onsager-Machlup functional \cite{om1978, om_functional_pinski}. 
\begin{remark}
  The introduction of Onsager-Machlup functional and Stratonovich integral is
  not necessary. However, it is helpful to allow us to follow the presentation of
  \cite{ce_tutorial}, \cite{ce_book} and to compare with the case of random
  variables.
\end{remark}
\begin{remark}
  \label{remark-general}
  The Onsager-Machlup functional can also be generalized to the following dynamics.
  \begin{enumerate}
    \item
      SDEs with non-constant diffusion coefficients. We consider 
      \begin{align}
	\vspace{-9mm}
    \begin{split}
      d z_s &= b(z_s) ds + \sqrt{2\beta^{-1}} \sigma(z_s)\,dw_s, \quad 0 < s < \tau \\
      z_0&=x
    \end{split}
    \label{dynamics-1-sigma}
    where $a(z) = \sigma(z)\sigma(z)^T$ is $n \times n$ positive definite matrix. We have 
    $d\mu = \exp(-\frac{\beta}{2} I_1(z)) d\nu$, with 
    \begin{align*}
      I_1(z) = \frac{1}{2} \int_0^\tau \Big|\frac{dz}{ds} -
      b(z_s)\Big|_{a(z)}^2 ds
    \end{align*}
     where Ito integral is used and $|v|_a = \langle v, a^{-1}v\rangle$,
     $\forall v \in \mathbb{R}^n$.
    \item
      System contains both stochastic and deterministic equations. For
      simplicity, here we consider dynamics 
      \begin{align}
      \begin{split}
	dx_s & = b_1(x_s, y_s)\, ds \\
	dy_s & = b_2(x_s, y_s)\, ds + \sqrt{2\beta^{-1}} dw_s
      \end{split}
      \label{ode-sde}
      where $x_s \in \mathbb{R}^k, y_s, w_s \in \mathbb{R}^{n-k}$, $b_1 :
      \mathbb{R}^k \times \mathbb{R}^{n-k} \rightarrow \mathbb{R}^k$, $b_2 :
      \mathbb{R}^k \times \mathbb{R}^{n-k} \rightarrow
      \mathbb{R}^{n-k}$. We can still assign a probability measure $\mu$ on the
      \textbf{trajectory space}, i.e. $\forall z = (x,y) $ is a trajectory, we have 
      $d\mu = \exp(-\frac{\beta}{2} I_1(z)) d\nu$, where $\nu$ is Wiener
      measure on $\mathbb{R}^{n-k}$ and 
      \begin{align*}
	I_1(z) = \frac{1}{2} \int_0^\tau \Big|\frac{dy}{ds} - b_2(x_s, y_s)
	\Big|^2 ds\,.
      \end{align*}
      Notice that only the stochastic equations are used.
  \end{enumerate}
\end{remark}
\section{Cross-entropy method for importance sampling}
\label{sec-ce-is}
In this section, we discuss how to apply the cross-entropy
method \cite{ce_tutorial, ce_book} to guide the
importance sampling in diffusion process.

Let $\mathcal{D}$ be certain subset in the path space and
$\textbf{1}_{\mathcal{D}}$ be the
characteristic function on $\mathcal{D}$. 
In the following, for simplicity, we will use $z$ to denote a path trajectory
$\{z_s\}$ and keep the notation $z_s \in \mathbb{R}^n$ for the state at time $s$. 
Let $F(\cdot)$ be certain path functional and will be written as $F(z)$ for
path $z$. 
We consider Monte Carlo method with importance sampling to compute the quantity
\begin{align}
  l = \mathrm{E}(F(z)~|~z_0 = x)
  \label{mean-l}
\end{align}
where we assume $F(\cdot)$ has the following type
\begin{align}
  F(z) = \textbf{1}_{\mathcal{D}}(z) \quad \mbox{or} \quad F(z) = \exp(-\beta\int_0^\tau G(z_s) \,ds).
  \label{path-fun-f}
\end{align}
  A special and interesting case is when $\mathcal{D} = \{z, \tau(z) \le \lambda\}$,
  $\lambda > 0$ is fixed. Then $l = \mathrm{E}(\textbf{1}_{\mathcal{D}}(z)) =
  \mathcal{P}(\tau \le \lambda)$, i.e. the probability of the trajectories
  stop before time $\lambda$. This quantity may provide us some details of the
  distribution of the stopping time $\tau$ and help us understand the
  behaviors of the dynamics.

We suppose $m$ basis functions on state space $\{\phi_i\}_{1\le i \le m}$ are
available, where $\phi_i : \mathbb{R}^n\rightarrow \mathbb{R}$. 
We will add a control force to the original dynamics (\ref{dynamics-1})
and the importance sampling will be realized by simulating trajectories from
this modified dynamics. Specifically, given $\bm{\omega} = (\omega_1, \omega_2,
\cdots, \omega_m)^T \in \mathbb{R}^m$,
the corresponding control force is state-dependent and is supposed to be 
\begin{align}
  u(x; \bm{\omega}) = \sum\limits_{i = 1}^{m} \omega_i \nabla \phi_i(x), \quad x \in
  \mathbb{R}^n\,.
  \label{control-force}
\end{align}
Therefore, for fixed parameter $\bm{\omega}$, the controlled dynamics is 
\begin{align}
\begin{split}
  d z_s &= b(z_s) ds + \sqrt{2}u(z_s;\bm{\omega}) ds + \sqrt{2\beta^{-1}} dw_s, \quad t \le s \le \tau \\
     z_0 &=z
  \end{split}
\label{dynamics-2}
\end{align}
Let $\mu_{\bm{\omega}}$ be the probability measure of dynamics (\ref{dynamics-2}). 
From Section \ref{sec-om-functional},  we know $\mu_{\bm{\omega}}$ has density 
$f(z;\bm{\omega}) = \exp(-\frac{\beta}{2} I(z ; \bm{\omega}))$ w.r.t. Wiener
measure $\nu$, with
\begin{align}
  I(z ; \bm{\omega}) = \frac{1}{2}\int_0^\tau \Big|\frac{dz}{ds}- b(z_s) -
  \sqrt{2}\sum\limits_{i = 1}^{m} \omega_i \nabla \phi_i(z_s)\Big|^2 ds + \beta^{-1}
  \int_0^\tau \operatorname{div}\, \Big(b(z_s) + \sqrt{2}\sum\limits_{i =
  1}^{m} \omega_i \nabla \phi_i(z_s)\Big) ds
  \label{om-fun-w}
\end{align}
and the original dynamics (\ref{dynamics-1}) corresponds to $\bm{\omega} = \bm{0}$ (we use $\mu = \mu_{\bm{0}}$).
Given a probability measure $\mu_g$ on path space with $d\mu_g = g(z)d\nu$, 
we can design a unbiased Monte Carlo estimator for (\ref{mean-l}) as 
\begin{align}
  l_N = \frac{1}{N} \sum_{i=1}^{N}  \frac{F(z_i)f(z_i;\bm{0})}{g(z_i)}
\end{align}
where trajectories $z_i, i = 1, \cdots, N$ are sampled from measure $\mu_g$.
The optimal measure $\mu_{g^*}$, which gives
zero-variance estimator, should satisfy 
\begin{align}
  g^*(z) = \frac{F(z)f(z;\bm{0})}{l}, \hspace{3mm} \forall z
  \label{optimal_pdf}
\end{align}
However, $g^*(z)$ depends on the path $z$ and is not easy to compute.

The idea of cross-entropy method is to find the best approximation of $g^*(z)$ among
$f(z;\bm{\omega}), \bm{\omega} \in \mathbb{R}^m$. More specifically, define the
Kullback-Leibler divergence between
two measures $d\mu_1 = g_1d\nu$, $d\mu_2 = g_2d\nu$ as
\begin{align}
  D(\mu_1, \mu_2) = \mathrm{E}_{\mu_1} \log \frac{d\mu_1}{d\mu_2} = \int
  g_1\log g_1\, d\nu - \int g_1\log g_2\,d\nu,
  \label{kl-dist}
and we try to minimize
\begin{align}
  \min_{\bm{\omega} \in \mathbb{R}^m} D(\mu_{g^*}, \mu_{\bm{\omega}}).
  \label{kl-dist-d}
Using (\ref{optimal_pdf}) and (\ref{kl-dist}), it is equivalent to solve
\begin{align}
  \max_{\bm{\omega} \in \mathbb{R}^m} \mathrm{E}_{\mu} ( F(z) \log
  f(z;\bm{\omega}))
  = \max_{\bm{\omega} \in \mathbb{R}^m} \mathrm{E}_{\mu_{\bm{v}}} ( F(z)
  \eta(z;\bm{v}) \log f(z;\bm{\omega})), \hspace{3mm} \forall \bm{v} \in
  \mathbb{R}^m
  \label{ce-2}
where
\begin{align}
  \eta(z;\bm{v}) = \frac{f(z;\bm{0})}{f(z;\bm{v})}.
For numerical purpose, we consider the discrete formulation of (\ref{ce-2})
\begin{align}
\max_{\bm{\omega} \in \mathbb{R}^m} \frac{1}{N} \sum_{i=1}^{N} F(z_i)
\eta(z_i; \bm{v}) \log f(z_i;\bm{\omega}) 
\label{ce-2-discrete}
where $z_i, i = 1, \cdots, N$ are trajectories generated from $\mu_{\bm{v}}$. 
Taking gradient, the maximizer of (\ref{ce-2-discrete}) could be found by solving
\begin{align}
  \frac{1}{N} \sum_{i=1}^{N} F(z_i) \eta(z_i; \bm{v}) \nabla_{\bm{\omega}} \log
  f(z_i;\bm{\omega}) = 0.
  \label{max-gradient}
With formula (\ref{om-fun-w}), equations (\ref{max-gradient}) could be written more explicitly
as linear equations $A\bm{\omega} = \bm{r}$, where $\forall 1\le i,j\le m$, 
\begin{align}
  \begin{split}
  A_{ij} &= \frac{1}{N}\sum_{k = 1}^{N} F(z_k)\eta(z_k; \bm{v}) \int_0^\tau
  \nabla\phi_i(z_{k,s})\nabla\phi_j(z_{k,s})\,ds, \\
     r_i &= \frac{1}{\sqrt{2}N}\sum_{k = 1}^{N} F(z_k)\eta(z_k; \bm{v})
  \Big[\int_0^\tau \Big(\frac{dz_k}{ds} - b(z_s)\Big)\nabla\phi_i(z_{k,s})\,ds
  - \beta^{-1} \int_0^\tau \operatorname{div} \nabla \phi_i(z_{k,s})\,ds\Big]
  \end{split}
  \label{linear-eqn}
Notice that (\ref{max-gradient}) holds $\forall \bm{v} \in \mathbb{R}^m$ and

>>>>

\newcommand{\rates}[2]{\displaystyle \mathrel{\longrightleftharpoons^{#1\mathstrut}_{#2}}}
\newcommand{\cL}{{\mathcal L}}
\newcommand{\cI}{{\mathcal I}}
\newcommand{\cF}{{\mathcal F}}
\newcommand{\cD}{{\mathcal D}}
\newcommand{\bR}{{\mathbb R}}
\newcommand{\R}{{\mathbb R}}
\newcommand{\Q}{{\mathbb Q}}
\newcommand{\N}{{\mathbb N}}
\newcommand{\bE}{{\mathbf E}}
\newcommand{\bP}{{\mathbf P}}

\newcommand{\eps}{\epsilon}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\argmin}{\mathop{\rm argmin}}%
\newcommand{\bk}[2]{\left\langle #1,#2\right\rangle}
\newcommand{\ip}{\langle\cdot,\cdot\rangle}
\newcommand{\di}{{\rm div}\,}
\newcommand{\vol}{{\rm vol}}
\newcommand{\wrt}{with respect to }

@@

\begin{document}
\title{Applications of the cross-entropy method to importance sampling and optimal control of diffusions}
\author{Wei Zhang\textsuperscript{2}\and Han Wang\textsuperscript{1} \and
  Carsten Hartmann\textsuperscript{1,*}\and 
Marcus Weber\textsuperscript{2}\and Christof Sch\"{u}tte\textsuperscript{1,2}}
\footnotetext[1]{Institute of Mathematics, Freie Universit\"{a}t Berlin, Arnimallee 6, 14195 Berlin, Germany}
\footnotetext[2]{Zuse Institute Berlin, Takustrasse 7, 14195 Berlin, Germany}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[1]{Corresponding author. Email: chartman@mi.fu-berlin.de}
\date{}
\maketitle
\begin{abstract}
 We study the cross-entropy method for diffusions. One of the results is a
  versatile cross-entropy algorithm that can be applied to the design of efficient importance sampling strategies for rare events as well as optimal
  control problems. The approach is based on the minimization of a suitable cross-entropy functional, based on a parametric family of exponentially tilted probability distributions. We illustrate the new algorithm with several numerical examples and discuss algorithmic issues and possible extensions of the method.
\end{abstract}
\begin{keywords}
importance sampling, optimal control, cross-entropy method, rare events, change of measure. 
\end{keywords}
\begin{AMS}
\end{AMS}

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{intro}
This article deals with the application of the cross-entropy method to diffusion processes, specifically, with the application to  
%Diffusion processes have been extensively investigated because of their wide range of applications in mathematics, physics, economics, and other fields. 
importance sampling for rare events and optimal control. Generally, the cross-entropy method is a Monte-Carlo method that was originally developed for the efficient simulation of rare events in queuing models and that has, in the meantime, been extended to, e.g., combinatorial optimization or analysis of networks \cite{ce_book,ce_tutorial}. To our knowledge, however, the cross-entropy approach has not been analysed or used in connection with diffusion processes, even though there have been significant research activities in the direction of efficient algorithms for importance sampling and optimal control high dimensional multiscale diffusions; see, e.g., \cite{ip-dupuis-multiscale, ip-eric,
zhws13} for some ideas related to importance sampling of rare events or \cite{control_schuette, zlph2013} for problems in 
optimal control.

%Importance sampling is a quite general variance reduction technique 
%for Monte Carlo simulations that is essential in rare event simulation, especially if applied to Monte Carlo simulations in path space. 
%Seemingly different, optimal control aims at solving the problem of how to driven the dynamical system in certain
%optimal way and has strong applicational backgrounds in finance, climate
%science, mechanics \cite{fleming2006}.

We will exploit the fundamental duality between importance sampling and optimal control, which arises due to the fact that both problems admit a variational formulation that boils down to finding an optimal transformation of the underlying path space measure \cite{WangDupuis2004}. Algorithms for computing an optimal change of measure problems usually seek an approximation of the optimal measure \wrt some distance on the space of (probability) measures. Here, we will use the Kullback-Leibler 
divergence, which, although not a metric, is a numerically convenient and widely used similarity measure for probability measures. 
The cross-entropy method then provides a general algorithm to find the minimizer of the
Kullback-Leibler divergence among a family of parameterized
probability measures, and the main purpose of this paper is to formulate the 
method in the context of diffusions, i.e.~for probability measures on the space of continuous paths, and discuss 
its application to importance sampling and optimal control.

This paper is organized as follows. The cross-entropy method in path space is
outlined in Section~\ref{sec-ce-is} and discussed in the context of importance
sampling and the dual optimal control problem. Section~\ref{sec-ce-algo} is
devoted to the formulation of the cross-entropy algorithm for diffusions. Several numerical examples are studied in Section~\ref{sec-examples}. We summarize our findings in Section~\ref{sec-discuss}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The cross-entropy method in path space}
\label{sec-ce-is}
In this section, we discuss how to use the cross-entropy
method for stochastic differential equations. The mathematical set-up will be largely based on the application of the method to importance sampling, which is the most standard application of the cross-entropy method in the literature. The associated (dual) optimal control problem will be briefly discussed at the end of this section.

\subsection{Problem set-up}

Consider  $z_s \in \mathbb{R}^n$ satisfying 
\begin{equation}\label{dynamics-1}
\begin{aligned}
  d z_s & =  b(z_s) ds + \sqrt{2\eps }\, dw_s\,, \quad 0\le s\le T\\
  z_{0} & = x
\end{aligned}
\end{equation}
where $\eps>0$ is constant, $b\colon \mathbb{R}^n \to \mathbb{R}^n$ is a smooth, locally Lipschitz-continuous vector field, and $w$ is 
$n$-dimensional Brownian motion. Further let $O \subset \mathbb{R}^n$ be open and bounded and call 
\begin{equation}\label{tau}
\tau = \inf\{s>0\colon \, z_s \notin O\}  
\end{equation}
the first exit time of the set $O\subset\mathbb{R}^n$.
%
%Let $\mathcal{D}$ be certain subset in the path space and
%$\textbf{1}_{\mathcal{D}}$ be the
%characteristic function on $\mathcal{D}$. 
In the following we will use $Z$ to denote a path trajectory
$\{z_s\}$ and use the notation $z_s \in \mathbb{R}^n$ for the state at time $s$. 
Accordingly we denote by $F(Z)$ a path functional that, throughout this paper, is assumed to be of the form
\begin{equation}
F(Z) = \exp\left(-\frac{1}{\eps}\int_0^{\tau\wedge T} G(z_s) \,ds - \frac{1}{\eps} H(z_{\tau\wedge T})\right).
  \label{path-fun-f}
\end{equation}
for some continous and bounded functions $G,H\colon\R^{n}\to\R$ and with $\tau\wedge T=\min\{\tau,T\}$. We consider a Monte Carlo method to compute the quantity
\begin{align}
  \ell(x) =\bE(F(Z)),
  \label{mean-l}
\end{align}
with $\bE(\cdot) = \bE(\cdot|\,z_0 = x)$ denoting the conditional expectation over all realizations of (\ref{dynamics-1}) starting at $z_{0}=x$.  
A special and interesting case is when $G = 0$ and $H=-\eps\log{\mathbf 1}_{O^{c}}$ with $O^{c}=\R^{n}\setminus O$ denoting the complement of the set $O$, in which case 
\begin{align}
  \bE({\mathbf 1}_{O^{c}}(z_{\tau\wedge T})) = \bP(\tau\le T),
  \label{mean-chi}
\end{align}
is the probability of trajectories starting at $z_{0}=x$ to reach the boundary of $O$ before time $T$. In many cases, this quantity provides details of the distribution of the stopping time $\tau$ and can be used to understand the behaviour of the dynamics.

%%%%%%%%%%%%%%%%%%
\subsection{Cross-entropy method for importance sampling}

We now formulate the cross-entropy method in path space following the relevant literature \cite{ce_tutorial,ce_book}. To this end we first introduce  the general concept of importance sampling in path space: Suppose we are able to generate path samples from a family of probability
measures $\{\mu_{\blambda}\}_{\blambda \in
\cF}$ on the space of continuous functions $C([0,T],\R^{n})$ that are
parametrized by $\blambda \in \cF\subset \mathbb{R}^m$ where the  dynamics
(\ref{dynamics-1}) corresponds to $\blambda = \bm{0}$; for the sake of
simplicity,  we set $\cF=\mathbb{R}^m$. We use the shorthand $\mu = \mu_{\bm{0}}$ and refer to $\mu_{\blambda\neq\bm{0}}$ as the \emph{tilted probability measure}. We further assume that every $\mu_{\blambda}$ has a probability density $f(\cdot;\blambda)$ \wrt the scaled Wiener measure $\nu_{\eps}$.\footnote{The scaled measure $\nu_{\eps}$ is the probability measure induced by the scaled Brownian motion $\sqrt{2\eps} w_{s}$ on the space $C([0,T],\R^{n})$. It is related to the standard Wiener measure underlying the standard Brownian motion $w_{s}$ by rescaling of time, which follows from the fact that $w_{s}$ and $\sqrt{2\eps}\,w_{s/(2\eps)}$ have the same law.}

\subsubsection*{Importance sampling.}
%
The idea of importance sampling is---instead of drawing samples from the measure $\mu$---to generate samples from an alternative probability measure $\eta = g(\cdot)\nu_{\eps}$ that is absolutely continuous \wrt $\mu$, but that yields better Monte-Carlo estimators, e.g., having smaller variance or smaller relative error \cite{Liu2008}. Using independent draws from $\eta$ an unbiased Monte Carlo estimator of (\ref{mean-l}) is then given by 
\begin{align}
  \ell_N = \frac{1}{N} \sum_{i=1}^{N}  \frac{F(\tilde{Z}_i)f(\tilde{Z}_i;\bm{0})}{g(\tilde{Z}_i)}\,,
\end{align}
where the trajectories $\tilde{Z}_i, i = 1, \cdots, N$ are independent realizations from $\eta$. It is well known that the optimal measure $\eta^*$ that minimizes the variance of the estimator has the density
\begin{align}
  g^*(Z) = \frac{F(Z)f(Z;\bm{0})}{\ell}\,.
  \label{optimal-pdf}
\end{align}
It is easy to see that the thus defined $\eta^{*}$ yields a zero variance estimator, note however that it depends on $\ell=\bE(F(Z))$, which is the quantity that we want to compute.

The idea of the cross-entropy method is to find the best approximation of $\eta^*$ among the family
$\mu_{\blambda}, \blambda \in \cF$ of tilted probability measures. The approach is based on minimizing the 
Kullback-Leibler divergence, which in our case can be defined as follows:
given two probability measures $\mu_1 = g_1\nu_{\eps}$, $\mu_2 =
g_2\nu_{\eps}$ that are absolutely continuous \wrt the scaled Wiener measure, the \emph{Kullback-Leibler divergence} or \emph{relative entropy} between $\mu_{1}$ and $\mu_{2}$ is defined as
\begin{align}
  D(\mu_1, \mu_2) = \bE_{\mu_1}\!\left(\log \frac{d\mu_1}{d\mu_2}\right) 
  \label{kl-dist}
\end{align}
where the expectation \wrt the measure $\mu_{1}$ is defined as
\begin{align}
  \bE_{\mu_1}\!\left(\log \frac{d\mu_1}{d\mu_2}\right) = \int \log \frac{d\mu_1}{d\mu_2}\,d\mu_{1} = \int
  g_1\log g_1\, d\nu_{\eps} - \int g_1\log g_2\,d\nu_{\eps}\,.
  \label{kl-dist2}
\end{align}

\subsubsection*{Cross-entropy method I.}
%
The cross entropy method now seeks an optimal change of measure by solving the minimization task
\begin{align}
  \min_{\blambda\in\R^{m}} D(\eta^*, \mu_{\blambda}).
  \label{kl-dist-d}
\end{align}
for the tilt parameter $\blambda\in\R^{m}$. Not knowing what $\eta^{*}$ is, this still sounds like an infeasible minimization problem. It turns out, however, that we need to know $\eta^{*}$ only up to a constant prefactor, which in our case, since $x$ is fixed, eliminates the unknown quantity $\ell=\ell(x)$. Using (\ref{optimal-pdf}) and (\ref{kl-dist}), the minimization problem is equivalent to the following maximization problem
\begin{align}\label{ce-1}
  \max_{\blambda \in \bR^m} \bE_{\mu} ( F(Z) \log
  f(Z;\blambda))\,.
\end{align}

For the efficient numerical solution of (\ref{ce-1}) it is often convenient to allow for drawing samples from a probability measure that  somehow ``in between'' $\mu$ and $\mu_{\blambda^{*}}$. This will give us some extra freedom to use, e.g., an iterative solver for the maximization problem (\ref{ce-1}). Letting $\bm{v}\in \bR^{m}$  denote an arbitrary family parameter, our maximization problem reads 
\begin{align}
  \max_{\blambda \in \mathbb{R}^m} \bE_{\mu_{\bm{v}}} ( F(Z)
  h(Z;\bm{v}) \log f(Z;\blambda)),
  \label{ce-2}
\end{align}
where
\begin{align}
  h(Z;\bm{v}) = \frac{f(Z;\bm{0})}{f(Z;\bm{v})}.
\end{align}
An unbiased estimator of (\ref{ce-2}) is 
\begin{align}
\max_{\blambda \in \mathbb{R}^m} \frac{1}{N} \sum_{i=1}^{N} F(\tilde{Z}_i)
h(\tilde{Z}_i; \bm{v}) \log f(\tilde{Z}_i;\blambda) 
\label{ce-2-discrete}
\end{align}
where $\tilde{Z}_i, i = 1, \cdots, N$ are independent realizations generated from $\mu_{\bm{v}}$. The necessary condition for $\blambda^{*}$ being a maximizer of (\ref{ce-2-discrete}) is obtained by taking the gradient \wrt $\blambda$: 
\begin{align}
\sum_{i=1}^{N} F(\tilde{Z}_i) h(\tilde{Z}_i; \bm{v}) \nabla_{\blambda} \log
  f(\tilde{Z}_i;\blambda) = 0\,.
  \label{max-gradient}
\end{align}

The degree of difficulty when solving (\ref{max-gradient}) numerically of course depends on the parameterization of the tilted family of distributions that will be introduced in Section \ref{sec-ce-algo} below.

%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cross-entropy method for optimal control}
\label{subsec-ocp}
In this section, we consider applying the cross-entropy method to solve optimal control problems
for diffusion processes. 
%
To this end consider the optimal control problem with cost
function \cite{control_schuette,zlph2013}
\begin{equation}\label{cost}
  J(u) = \bE_{\mu_u}\left(\int_0^{\tau\wedge T} G(z_{s}) + \frac{1}{4} |u_{s}|^2
ds  + H(z_{\tau\wedge T})\right),
\end{equation}
with bounded continuous function $G,H\colon\R^{n}\to\R$ and $u_{s}\in\R^{n}$ being a measurable control that will specified below. The expectation $\bE_{\mu_u}$ \wrt the probability measure $\mu_u$ is understood as the expectation over all realizations of the controlled dynamics
\begin{equation}
 \begin{aligned}
   d z_s &= \left(b(z_s) +  u_{s}\right)ds + \sqrt{2\eps }\,dw_s, \quad 0 \le s \le T \\
    z_0&=x
  \end{aligned}
  \label{ctr-sde}
\end{equation}
We suppose that $G\ge 0$ and, for the ease of notation, we set $H=0$. We wish to minimize (\ref{cost}) under the constraint (\ref{ctr-sde}) and over all measurable controls $u$ that are adapted to the filtration generated by the Brownian motion driving the dynamics (\ref{ctr-sde}). Then, given suitable conditions on the vector field $b$, it is known that this optimal control problem has a unique solution in form of a Markovian feedback control. That is, for some continuous and bounded function $c\colon[0,T]\times \R^{n}\to\R^{n}$, it holds that 
\begin{equation}\label{feedback}
 \hat{u}_{s} = c(s, z_{s})\,.
\end{equation}
Now call $\mu, \hat{\mu}$ the probability measures on the path space  $C([0,T],\R^{n})$ corresponding to (\ref{ctr-sde}) with $u = 0$ and $\hat{u}$. Then, using the Legendre-type dual relation, we have \cite{DaiPra1996,HaEtal14} 
    \begin{align}
      J(\hat{u}) = -\eps \log \bE_\mu\left(\exp\left(-\frac{1}{\eps}\int_0^{\tau\wedge T} G(z_s) \,ds\right) \right) ,
      \label{dual-relation}
    \end{align}
    where, by Jensen's inequality (see \cite[Sec.~VI.2]{fleming2006}), we know that $\hat{\mu}$-a.s.
    \begin{align}
    \exp\left(-\frac{1}{\eps}\int_0^{\tau\wedge T} G(z_s)\,ds\right)\frac{d\mu}{d\hat{\mu}} =
    \bE_\mu\left(\exp\left(-\frac{1}{\eps}\int_0^{\tau\wedge T} G(z_s)\, ds\right)\right) 
    \label{jensen-as}
  \end{align}
From the above we conclude that (see \cite[Sec.~VI.3]{fleming2006} for details)
  \begin{equation}\label{cost-2}
  \begin{aligned}
    J(u) &= \bE_{\hat{\mu}}\!\left(\int_0^{\tau\wedge T} \left(G(z_s) + \frac{1}{4} |u_{s}|^2\right)\,ds\right)\frac{d\mu_u}{d\hat{\mu}}  \\
    & = J(\hat{u}) + \bE_{\hat{\mu}}\!\left(\left(\eps \log
    \frac{d\mu}{d\hat{\mu}} + \frac{1}{4} \int_0^{\tau\wedge T} |u_{s}|^2\,ds\right)\frac{d\mu_u}{d\hat{\mu}}\right)
\end{aligned}
\end{equation}
After some rearrangement and simplification, we get the following simple relationship: 
\begin{align}
J(u) = J(\hat{u}) + \eps D(\mu_u, \hat{\mu}).
\label{J-entropy}
\end{align}
where $D(\cdot, \cdot)$ is the Kullback-Leibler divergence as defined in (\ref{kl-dist}).
